Eagle: multi-locus association mapping on a genome-wide scale made routine

### 2019-02-08 bioinformaticspaper.tex
Just submitted to Bioinformatics

\documentclass{nature}
\usepackage{xr}
\bibliographystyle{naturemag}
\renewcommand{\baselinestretch}{1.3} 
\usepackage{graphicx}
\usepackage{bm}
\usepackage{lineno}
\linenumbers

@@

\title{Eagle makes multi-locus association mapping on a genome-wide scale routine}
\author{Andrew W. George$^{1,*}$, Arunas Verbyla$^1$, Joshua Bowden$^2$}

@@

\begin{affiliations}
\item Data61, CSIRO, Australia.
\item IM \&T, CSIRO, Australia.
\end{affiliations}

\begin{abstract}
We present Eagle, a new method for multi-locus association mapping. The motivation for developing Eagle was to make multi-locus association mapping "easy" and the method-of-choice. Eagle's strengths are that it a. is considerably more powerful than single-locus association mapping b. doesn't suffer from multiple testing issues c. is threshold free d. gives results that are immediately interpretable and e. has a computational footprint comparable to single-locus association mapping. By conducting a large simulation study, we will show that Eagle finds true and avoids false SNP-trait associations better than competing single- and multi-locus methods. We also analyse data from a published mouse study. Eagle found over two times more validated findings than the state-of-the-art single-locus method. Eagle has been implemented as an R package, with a web-based Graphical User Interface (GUI) for users less familiar with R.
\end{abstract}

Over the past decade,  genome-wide association studies (GWASs) have changed considerably in both their analysis and design. Early studies
 followed a case-control design. Association mapping methods were no more complicated than contingency table tests or simple 
linear regression. These designs though had a tendency to yield spurious findings if there was unrecognised population stratification 
\cite{cardon2003population}. This prompted a shift towards family-based designs and score tests, such as the transmission/disequilibrium test (TDT)  and its variants \cite{spielman1996tdt}. Today, instead of by design, it is through statistical modelling that we account for the effects of population stratification \cite{price2010new}. This has meant that data can be collected from general populations, even if these populations are highly structured. Analysis via sophisticated association mapping methods, such as linear mixed model based approaches,  is now almost routine \cite{purcell2007plink, bradbury2007tassel}.

What has not changed is that it remains common practice to analyse genome-wide association study (GWAS) data on a locus-by-locus basis. This is despite there being several significant problems with analysing data in this way. 
First, the aim of association mapping is to identify regions of the genome that house genes that are influencing a trait. 
The identification of these regions from these analyses is not always straightforward. GWAS results are reported, typically, via Manhattan plots 
that plot the -log$_{10}$ of the $p$ value for each locus against the map position of the locus. The $p$ value is obtained by testing the statistical 
significance of a single nucleotide polymorphisms (SNP) when treated as an effect in an appropriate model. 
The location of peaks in this plot identify genomic 
regions of interest. Inferring the exact number of regions though can be difficult if the peaks are not well separated. Second, when multiple statistical tests are performed, the probability of wrongly accepting a result (type 1 error) is inflated. This is known as the multiple testing problem. 
Many different solutions have been offered \cite{storey2003statistical, li2005adjusting, de2005efficiency}. Yet, there is still no well accepted way of correcting for multiple testing in the context of genome-wide association mapping. Third, many of the traits whose genetic secrets we are trying to discover are complex. There will be multiple SNPs in linkage disequilibrium with genes that are influencing the trait. Yet, a locus-by-locus mapping approach only assesses the evidence for association between a single marker locus and trait.

It is somewhat surprising then that multi-locus association mapping methods haven't attracted more attention. Methods based on 
regularisation techniques, such as ridge regression \cite{shen2013novel}  and lasso \cite{rakitsch2013lasso}, measure all locus-trait associations simultaneously. 
Here, multiple testing is not an issue. These techniques though are computationally demanding. Also, their results can be difficult to interpret. The strength of association is not measured by a $p$ value but by the size of the regression coefficient for the SNP in the model. More recently, associations have started to be mapped with random forests \cite{szymczak2016r2vim}. Similar to regularisation techniques though, it is not clear how to infer genomic regions of interest from their findings. A multi-locus method that does show promise is the multiple-locus linear mixed model method \cite{segura2012efficient}. The best multi-locus model is built with simple forward selection. Results are immediately interpretable but here, computation becomes challenging for large datasets.

@@

@@

We compared Eagle, in terms of computational and statistical performance, against seven other association mapping methods. 
We chose methods that almost all had been purpose built for genome-wide analysis, that could handle data from quantitative traits, and where the methods had been implemented in freely available computer programs or packages. Two of the methods are based on single-locus (or locus-by-locus) models and five are based on multi-locus models. Of the many ways of performing single-locus association mapping, we chose 
GEMMA \cite{zhou2012genome}  and FaST-LMM \cite{lippert2011fast} because of their popularity and computational speed. 
For multi-locus association mapping, we chose bigRR \cite{shen2013novel}, glmnet \cite{Friedman2010glmnet}, 
LMM-Lasso \cite{rakitsch2013lasso}, MLMM \cite{segura2012efficient} , and r2VIM \cite{szymczak2016r2vim}.  
Each takes a different approach to multi-locus association mapping. A summary of the key attributes of the different computer programs/packages 
is given in Supplementary Table \ref{suptabsummary} (see Methods for further details).

@@

For each scenario, 100 replicates were generated. A single replicate consisted of SNP and quantitative trait data. 
Extra realism was introduced into the simulation study through the drawing of the SNP genotypes from the 1000 Genome Project, phase 3  \cite{10002010map}.
 The quantitative trait data were generated 
by selecting, randomly, a set of SNPs and assigning these loci additive allelic effects.  Random errors were then drawn from a normal distribution 
with variance set to give a heritability of 50\% for the trait. 
For each individual, a quantitative trait value was obtained by summing its random error and additive allelic effects. 
The number of randomly selected SNPs follows a Poisson distribution with mean 30. The size of the allelic effects 
 across the selected loci are equal.

@@

R2VIM is different from the rest in that it is a nonparametric approach for association mapping. It is based on random forests. Three important parameters needed to be  set. 
These were the number of trees, the number of variables for building a tree, and the minimum size of a terminal node. Ideally, these parameters would be "tuned" on a replicate-by-replicate 
basis \cite{boulesteix2012overview}. However, this was not practical here. We instead used the same settings as in \cite{szymczak2016r2vim} where 
the number of trees was set to 1000, the number of variables was set to 20\% of the number of SNPs, and 
  the minimum size of a node was set to 10\% of the sample size.
A relative importance measure was calculated 
for each SNP measuring its strength of association with the trait.

@@

\begin{figure}
\caption{Power verse false discovery rates for Eagle and the single-locus methods GEMMA and FaST-LMM. 
FaST-LMM was run where all the SNP data are used 
to estimate the relationship matrix (FaST-LMM$^{all}$)   and where genotype data from every five-hundredth SNP are used to 
estimate the relationship matrix (FaST-LMM$^{few}$). Eagle has substantially higher power than the single-locus methods.   }

\label{figpowersingle}
\begin{center}
\includegraphics[width=15cm, height=20cm]{power2main.eps}
\end{center}

@@

\begin{figure}
\caption{Median run times, in minutes, for the analysis of simulation study data from the six scenarios. 
Eagle is compared against five other multi-locus programs/packages (A) and two single-locus programs (B). 
The x- and y-axes are on a log scale for improved aesthetics. Eagle has the lowest run-times of the multi-locus 
programs/packages, sometimes by orders of magnitude. Eagle can even produce results faster than single-locus programs. 
The median run times for the programs/packages across the scenarios are given in the table (C). The entries in a blue font 
correspond to the lowest run-time for a scenario. 
 FaST-LMM$^{all}$ is where calculation of the similarity matrix is based on all the SNP data.  
 FaST-LMM$^{few}$ is where calculation  of the similarity matrix is based on a subset of the SNP data. }

%% plot formed via PowerPoint - ~/AM-Paper/Figure_power.pptx
%% script on mac
%% ~/AM-Paper/AM+/Plots_for_Paper/time_and_memory_plot_for_paper.R
\label{fig_time}
\centering
    \includegraphics[width=1.0\textwidth,natwidth=610,natheight=642]{Figure2_time.jpg}
\end{figure}

@@

We were interested in comparing results from Eagle with those from single-locus association mapping for a real data set.
 We chose to focus on data from a large outbred mouse study \cite{nicod2016genome}. This study was unusual in that it collected and analysed SNP dosages (continuous values from zero to one of expected allele counts)  instead of the more common SNP genotypes. Analyses based on dosages rather than discrete genotypes have been shown to have greater power for the detection of genes that are influencing a trait  \cite{zheng2011comparison}. By converting the dosages into genotypes and analysing the data with the single-locus program FaST-LMM, we obtained a subset of those findings reported in the original study. We then analysed the data with Eagle. Due to Eagle's increased power, we found SNP-trait associations not found with  FaST-LMM. However, we were 
 able to confirm the validity of these new findings as they matched what was found in the original study. Having the ability to confirm new findings  in a real study was 
 one of the primary motivators for choosing these data for analysis.

We repeated the single-locus analyses as first performed \cite{nicod2016genome}, except 
that we focused on autosomal SNPs and our analyses were based on SNP genotypes  rather than SNP dosages. 
In the original analysis, a genome-wide threshold that gave a false discovery rate of 5\%, was found via permutation. We followed the 
same empirical procedure but increased the number of permutations from 100 to 500 for more accurate thresholds.

We ran Eagle in three ways. 
Eagle chooses the best model via the extended Bayesian information criteria (extBIC) \cite{chen2008extended}. 
  The conservativeness of the extBIC can be adjusted by a single regularisation parameter  $\gamma$ that ranges from zero to one. 
In the simulation study, this parameter was set to one, its most conservative and default setting. The mouse data were also analysed 
under this setting (Eagle$^{default}$).  An alternate  \cite{chen2008extended} , less conservative way of setting $\gamma$ is to  let $\gamma = 1 - \frac{1}{(2\kappa)}$ with $\kappa=\frac{log(L)}{log(n_g)}$ where 
 $L$ is the number of loci that span the genome, and $n_g$ is the number of individuals/groups/lines/strains in the study
  (Eagle$^{alt}$). However, our preferred way is to set the $\gamma$ parameter for each trait via permutation (Eagle$^{optimal}$).  We used 100 permutations to set $\gamma$ to give a false  positive rate of 5\%.  This only took six times as long as a single analysis of the data. This is because the marker data need only be read  once, 
 and only the trait data changes across permutations leading to other computational efficiencies.  
  This permutation method has been implemented within the
 Eagle package.

@@

\begin{figure}
\caption{Genome-wide association mapping results from analyses of the mouse data for the single-locus method FaST-LMM and the 
multi-locus method
Eagle. Eagle was run under three settings; its default setting (Eagle$^{default}$),  an alternate less conservative setting based on 
the number of SNPs and sample size (Eagle$^{alt}$), and where the model selection had been optimised for a false positive rate 
of 5\% (Eagle$^{optimal}$).  The number of SNP-trait associations found are reported in the cells. }
\label{figmouse}
\begin{center}
%% obtained from bracewell by running ResultsV2.R from /data/geo047/MWAM/Mouse/OutBred/RScripts/
\includegraphics[width=15cm, height=15cm]{mouseresults.eps}
\end{center}

@@

\section{Discussion}
Eagle is a new linear mixed model based method (and R package) for multi-locus association mapping. It advances the state of association mapping in several ways. 
First, its computational footprint is much smaller than other multi-locus implementations. Eagle makes multi-locus analysis 
practical, even when the datasets are large. Second, the results from
 Eagle are immediately interpretable. They are the set of SNPs in strongest association with the trait where 
each SNP identifies a separate genomic region of interest. Third, it treats association mapping as a model selection problem, avoiding 
multiple testing issues and the need for significance thresholds. 
 As we saw in the simulation study, Eagle has considerably higher power than single-locus methods but is comparable in run time.
Also, when analysing the mouse data, Eagle found more than double the SNP-trait associations than 
with single-locus association mapping, the method of choice. Furthermore, these extra findings were all true.

>>>>

%%\documentclass{nature}
\documentclass{article}
\usepackage{xr}
\bibliographystyle{natbib}
\renewcommand{\baselinestretch}{1.4} 
\usepackage{graphicx}
\usepackage{bm}
\usepackage{lineno}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{authblk}

\linenumbers

@@

\title{Eagle: multi-locus association mapping on a genome-wide scale made routine}
\author[1]{Andrew W. George}
\author[2]{Arunas Verbyla}
\author[3]{Joshua Bowden}
\affil[1]{Data61, CSIRO, Brisbane, 4102, Australia.}
\affil[2]{Data61, CSIRO, Atherton, 4883, Australia.}
\affil[3]{IM \&T, CSIRO, Brisbane, 4067, Australia}
\date{}

@@

\begin{abstract} \noindent
\setstretch{1.6}{
{\bf Motivation:} We present Eagle, a new method for multi-locus association mapping. The motivation for developing Eagle was to make multi-locus association mapping "easy" and the method-of-choice. Eagle's strengths are that it a. is considerably more powerful than single-locus association mapping b. doesn't suffer from multiple testing issues c. is threshold free d. gives results that are immediately interpretable and e. has a computational footprint comparable to single-locus association mapping. \\
{\bf Results:} By conducting a large simulation study, we will show that Eagle finds true and avoids false SNP-trait associations better than competing single- and multi-locus methods. We also analyse data from a published mouse study. Eagle found over 100\% more validated findings than the state-of-the-art single-locus method.  \\
{\bf Availability and Implementation:} Eagle has been implemented as an R package, with a web-based Graphical User Interface (GUI) for users less familiar with R. It is freely available via the CRAN website  at https://cran.r-project.org.  \\
{\bf Contact:} andrew.george@csiro.au
}
\end{abstract}

\section{Introduction}
Over the past decade,  genome-wide association studies (GWASs) have changed considerably in both their analysis and design. Early studies
 followed a case-control design. Association mapping methods were no more complicated than contingency table tests or simple 
linear regression. These designs though had a tendency to yield spurious findings if there was unrecognised population stratification 
\citep{cardon2003population}. This prompted a shift towards family-based designs and score tests, such as the transmission/disequilibrium test (TDT)  and its variants \citep{spielman1996tdt}. Today, instead of by design, it is through statistical modelling that we account for the effects of population stratification \citep{price2010new}. This has meant that data can be collected from general populations, even if these populations are highly structured. Analysis via sophisticated association mapping methods, such as linear mixed model based approaches,  is now almost routine \citep{purcell2007plink, bradbury2007tassel}.

What has not changed is that it remains common practice to analyse genome-wide association study (GWAS) data on a locus-by-locus basis. This is despite there being several significant problems with analysing data in this way. 
First, the aim of association mapping is to identify regions of the genome that house genes that are influencing a trait. 
The identification of these regions from these analyses is not always straightforward. GWAS results are reported, typically, via Manhattan plots 
that plot the -log$_{10}$ of the $p$ value for each locus against the map position of the locus. The $p$ value is obtained by testing the statistical 
significance of a single nucleotide polymorphisms (SNP) when treated as an effect in an appropriate model. 
The location of peaks in this plot identify genomic 
regions of interest. Inferring the exact number of regions though can be difficult if the peaks are not well separated. Second, when multiple statistical tests are performed, the probability of wrongly accepting a result (type 1 error) is inflated. This is known as the multiple testing problem. 
Many different solutions have been offered \citep{storey2003statistical, li2005adjusting, de2005efficiency}. Yet, there is still no well accepted way of correcting for multiple testing in the context of genome-wide association mapping. Third, many of the traits whose genetic secrets we are trying to discover are complex. There will be multiple SNPs in linkage disequilibrium with genes that are influencing the trait. Yet, a locus-by-locus mapping approach only assesses the evidence for association between a single marker locus and trait.

It is somewhat surprising then that multi-locus association mapping methods haven't attracted more attention. Methods based on 
regularisation techniques, such as ridge regression \citep{shen2013novel}  and lasso \citep{rakitsch2013lasso}, measure all locus-trait associations simultaneously. 
Here, multiple testing is not an issue. These techniques though are computationally demanding. Also, their results can be difficult to interpret. The strength of association is not measured by a $p$ value but by the size of the regression coefficient for the SNP in the model. More recently, associations have started to be mapped with random forests \citep{szymczak2016r2vim}. Similar to regularisation techniques though, it is not clear how to infer genomic regions of interest from their findings. A multi-locus method that does show promise is the multiple-locus linear mixed model method \citep{segura2012efficient}. The best multi-locus model is built with simple forward selection. Results are immediately interpretable but here, computation becomes challenging for large datasets.

@@

The data were obtained from a large genome-wide association study that was performed in outbred mice \citep{nicod2016genome}. 
Phenotypic and genotypic data were available on 1,887 adult mice. 
The phenotypic data included raw and adjusted (for fixed effects) measurements from 200 behavioural, tissue, and physiological traits.  
Of these traits, 
43 yielded SNP-trait associations that could be corroborated through other independent published work. It was these 
43 traits that were the focus of our real data analyses. As in the original study  \citep{nicod2016genome}, our analyses 
were based on the adjusted traits.
Genotypic data were available on 359, 559 (353,697 autosomal) SNPs in the 
form of marker dosages (expected allele counts that ranged from zero to one). All missing data had been imputed. 
We converted the dosages into discrete genotypes 
by clustering around 0, 0.5, and 1, corresponding to SNP genotypes AA, AB, and BB, respectively. 
We focused our analyses on the autosomal SNPs.

Eagle is a method for multi-locus association mapping on a genome-wide scale. It is based on linear mixed models. It differs from most other single- and multi-locus association mapping methods. Eagle treats association mapping as a model selection instead of variable selection problem. Consequently,  we do not have to contend with multiple testing issues or having to construct significance thresholds.  
Eagle also 
reports as its findings only those SNPs that are in strongest linkage disequilibrium, and hence closest to the genes influencing a trait. 
The methodological foundation for Eagle comes from a whole-genome linkage analysis method that was developed for mapping 
quantitative trait loci in experimental crosses \citep{verbyla2007analysis}.

The specifics of the Eagle method are as follows. 
Eagle builds the "best" model iteratively, via forward selection. 
Suppose $s$ iterations of our model building process have already been performed. This means $s$ SNP-trait 
associations have been identified.  It also means that $s$ separate genomic regions of interest have been found.  
To perform the $s+1$th  iteration, we first fit the current model to the data. 
The (current) model is of the form 
\begin{equation}
\label{eq1}
\by = \bX \btau + \bZ \bu_g + \be
\end{equation}
where 
$\bX^{(n \; \times \; p)}$ and $\bZ^{( n \; \times \; n_g)}$ are known design matrices with $\bX$ being of full rank and $\bZ$ 
containing zeros and ones that assign the appropriate genetic effect to each measurement. 
The vector 
$\btau^{(p \; \times \; 1)}$ has $p$ fixed effects parameters including the intercept. The vector 
$\bu_g^{(n_g \; \times \; 1)}$ contains the 
genetic effects. The vector of residuals is 
$\be^{(n \; \times \;1)}$ whose distribution is assumed to follow $N(\bzero, \sigma^2_e \bI^{(n \; \times \; n)})$. 
So far,  this model differs little from standard linear mixed models for association mapping \citep{yu2006unified,zhao2007arabidopsis} 
However, 
it is how we specify $\bu_g$ that distinguishes our model from the others.

Second, we estimate the parameters of (\ref{eq1}) and (\ref{eq2}) via residual maximum likelihood (REML).  For complex models, REML
can be computationally demanding. However, our model only contains a single random effect ($\ba_{-S}$). Here, highly efficient single-dimension 
optimisation via spectral decomposition is possible \citep{kang2008efficient}.

Third, we identify the $(s+1)$th SNP that is in strongest association with the trait, based on the maximum score statistic
$t_j^2 = \frac{ \widetilde{a} _j^2}{\textrm{var}(\widetilde{a}_j)}$ where $\widetilde{a}_j$ is the best linear unbiased predictor (BLUP), 
and $\textrm{var}(\widetilde{a}_j)$ is its variance. This statistic is not only appealing intuitively, where we 
identify a SNP based on its (random) effect size and accuracy, but is justifiable, theoretically \citep{verbyla2012rwgaim}.

Fourth, we determine the importance of the $(s+1)$th selected SNP via a model selection strategy  \citep{verbyla2007analysis}. 
We begin by reforming (\ref{eq2}) where $S$ now contains the $s + 1$ selected SNP.  We then fit this new model to the data
via maximum likelihood and calculate its extended Bayesian information criteria (extBIC) \citep{chen2008extended}.  The 
extBIC is a model selection measure that takes into account the number of unknown parameters and the complexity 
of the model space.  It is well suited to the model selection problem in genome-wide association studies \citep{chen2008extended}. 
It is different to the model selection measure used in  \citep{verbyla2007analysis}.
If this new model has a larger extBIC than the current model, then the $s+1$th selected SNP is added to 
the current model and the above process is repeated. If this new model has a smaller extBIC than the current model, then the 
model building process is complete. The set of SNP in strongest association with the trait is the $s$ SNPs previously identified.

In practice, estimating the parameters of (\ref{eq2}) can be demanding, computationally. 
The vector $\ba_{-S}$ has $L-s$ random effects where in modern genome-wide association studies, 
$L$, the number of SNPs, can be extremely large.  An alternative model is given by 
Verbyla \citep{verbyla2012rwgaim,verbyla2014whole}. 
They show how to reformulate (\ref{eq2}) to be a model with a random effect with only $n$ elements
\begin{equation}
\label{eq3}
\bu_g = \sum_{k=1}^s  \bmm_{S_k} a_{S_k} + (\bM_{-S} \bM_{-S}^T)^{1/2} \ba^*_{-S}
\end{equation}
where $\ba^* \sim N(\bzero, \sigma_a^2 \bI^{(n_g \; \times \;  n_g)})$, and 
$(\bM_{-S} \bM_{-S}^T)^{1/2}$ can be calculated via singular value decomposition \citep{golub2012matrix}.  
Although it may not be obvious, the two models are equivalent, 
having identical variance structures. Yet, the computational cost of model (\ref{eq3}) compared to 
model (\ref{eq2}) is much less, due to the random term in model (\ref{eq3}) having only $n$ instead of $L-s$ 
effects needing estimating.

Verbyla \citep{verbyla2012rwgaim,verbyla2014whole} go on to show how to recover $\bwidetildea$ from estimates from model  (\ref{eq3})  with 
\begin{equation}
\bwidetildea = \left [ \bM_{-S}^T (\bM_{-S} \bM_{-S}^T)^{-1/2} \right ] \bwidetildea^*
\end{equation}
where its variance matrix is
\begin{equation}
\label{eq4}
\textrm{var}(\bwidetildea) = \bM_{-S}^T (\bM_{-S} \bM_{-S}^T)^{-1/2} \textrm{var}(\bwidetildea^*) (\bM_{-S} \bM_{-S}^T)^{-1/2} \bM_{-S}
\end{equation}
These values are needed in order to calculate the score statistic $t_j^2$ for identifying the SNP in strongest association with the trait. 
Fortunately, when calculating $t_j^2$, only the diagonal elements of the variance matrix are needed which simplifies the  calculation 
of (\ref{eq4}).

 \subsubsection{ Multi-locus methods:} We compare the computational and statistical performance of Eagle against five multi-locus methods. They are bigRR, LMM-Lasso, glmnet, 
MLMM, and r2VIM.  All but glmnet have been purposely designed for genome-wide association mapping. 
BigRR, LMM-Lasso, and glmnet are regression-based regularisation 
methods. BigRR is based on generalised ridge regression, LMM-Lasso is based on lasso, and glmnet is based on elastic net. 
Regularisation methods make parameter estimation possible in models  where the number of predictors is far greater than the number of samples. 
They allow the strength of association between all the SNPs and trait to be measured within a single model, simultaneously. 
A limitation of these methods though is that the statistical significance of the SNP effects cannot be easily determined. 
 Due to the adaptive nature of the estimation procedures, to do this 
analytically is challenging and is an area of active research \citep{lockhart2014significance}. Instead, we calculate significance empirically via 
stability selection (see below).

\subsubsection{Single-locus methods:} We also compare the performance of Eagle against two single-locus methods, GEMMA \citep{zhou2012genome} and FaST-LMM \citep{lippert2011fast}. Both are based on linear mixed models. The models have a single fixed effect for the SNP,  other fixed effects, 
a single random effect to account for familial relatedness (or polygenic background), and an error. The significance of the SNP effect in the model 
is a measure of the strength of association.  They are of the same computational complexity \citep{zhou2012genome}, and produce exact results.  
Both perform a single spectral decomposition of the relationship (or similarity) matrix $K$, use  an eigenvector matrix to rotate the data, 
and reformulate the (residual) log likelihood for easier computation. They do differ in their estimation procedure. GEMMA implements Newton-Raphson. 
FaST-LMM implements Brent's algorithm. Newton-Raphson is more complicated but has better convergence properties than Brent's algorithm. 
Both methods are state-of-the-art and have been implemented in highly efficient computer programs.

 Stability selection \citep{meinshausen2010stability}  is a subsampling strategy with a range of applications, including the estimation of the significance of effects in regression models where 
the number of parameters ($p$) can be much larger than the number of samples ($n$). We chose stability selection over permutation and other sampling procedures because of its low computational cost. We employe stability selection to estimate, empirically, the significance of the SNP effects from the regularisation method's analyses of the simulated data.

To obtain significance estimates via stability selection, we do the following. For LMM-Lasso and glmnet, we begin
 by performing a preliminary analysis of a replicate where  we tune the  regularisation parameter to yield 20 to 30 non-zero SNP effects.
This tuning of the regularisation parameter is done for data under each simulation scenario but it is not necessary to do it for every replicate within a scenario. It is sufficient to pick a single replicate at random upon which to tune the regularisation parameter. 
Here, we have the luxury of knowing the genetic conditions 
under which the simulated trait data are generated. We know that 20 to 30 SNP-trait associations is a reasonable number of findings to expect. 
 However, stability selection estimates are robust  \citep{meinshausen2010stability} 
 and the regularisation parameter can be tuned to any reasonable number of 
 non-zero effects. We sample repeatedly, without replacement, from the replicate. We draw 100 replicate subsets of size $n/2$.  
Each replicate subset is then analysed with LMM-Lasso and glmnet, with their regularisation parameter set to the value found in the preliminary analysis. 
A (probability) estimate of the statistical significance of a SNP effect is obtained by counting the 
number of times the SNP have a non-zero effect size over all the replicates divided by the number of replicates (which was 100).

The data are generated via data perturbation \citep{zhao2007arabidopsis}. 
Data perturbation amalgamates real with simulated data to generate replicates. 
It is a way of introducing greater realism into a simulation study. 
 Here, the genotype data are real, the 
quantitative trait data are simulated. 
 The SNP genotypes are drawn, 
according to the specifications of the scenario, from data collected from the 1000 Genome Project, version 3   \citep{10002010map}. 
Across scenarios, the SNP data differs. Across replicates within a scenario, the SNP data are the same.

@@

We compared Eagle, in terms of computational and statistical performance, against seven other association mapping methods. 
We chose methods that almost all had been purpose built for genome-wide analysis, that could handle data from quantitative traits, and where the methods had been implemented in freely available computer programs or packages. Two of the methods are based on single-locus (or locus-by-locus) models and five are based on multi-locus models. Of the many ways of performing single-locus association mapping, we chose 
GEMMA \citep{zhou2012genome}  and FaST-LMM \citep{lippert2011fast} because of their popularity and computational speed. 
For multi-locus association mapping, we chose bigRR \citep{shen2013novel}, glmnet \citep{Friedman2010glmnet}, 
LMM-Lasso \citep{rakitsch2013lasso}, MLMM \citep{segura2012efficient} , and r2VIM \citep{szymczak2016r2vim}.  
Each takes a different approach to multi-locus association mapping. A summary of the key attributes of the different computer programs/packages 
is given in Supplementary Table \ref{suptabsummary} (see Methods for further details).

@@

For each scenario, 100 replicates were generated. A single replicate consisted of SNP and quantitative trait data. 
Extra realism was introduced into the simulation study through the drawing of the SNP genotypes from the 1000 Genome Project, phase 3  \citep{10002010map}.
 The quantitative trait data were generated 
by selecting, randomly, a set of SNPs and assigning these loci additive allelic effects.  Random errors were then drawn from a normal distribution 
with variance set to give a heritability of 50\% for the trait. 
For each individual, a quantitative trait value was obtained by summing its random error and additive allelic effects. 
The number of randomly selected SNPs follows a Poisson distribution with mean 30. The size of the allelic effects 
 across the selected loci are equal.

@@

R2VIM is different from the rest in that it is a nonparametric approach for association mapping. It is based on random forests. Three important parameters needed to be  set. 
These were the number of trees, the number of variables for building a tree, and the minimum size of a terminal node. Ideally, these parameters would be "tuned" on a replicate-by-replicate 
basis \citep{boulesteix2012overview}. However, this was not practical here. We instead used the same settings as in \citep{szymczak2016r2vim} where 
the number of trees was set to 1000, the number of variables was set to 20\% of the number of SNPs, and 
  the minimum size of a node was set to 10\% of the sample size.
A relative importance measure was calculated 
for each SNP measuring its strength of association with the trait.

@@

\begin{figure}
\caption{
\setstretch{1.6}{
Power verse false discovery rates for Eagle and the single-locus methods GEMMA and FaST-LMM. 
FaST-LMM was run where all the SNP data are used 
to estimate the relationship matrix (FaST-LMM$^{all}$)   and where genotype data from every five-hundredth SNP are used to 
estimate the relationship matrix (FaST-LMM$^{few}$). Eagle has substantially higher power than the single-locus methods.   }
}
\vspace{-3cm}
\label{figpowersingle}
%%\begin{center}
\includegraphics[width=15cm, height=20cm]{power2main.eps}
%%\end{center}

@@

\begin{figure}
\caption{
\setstretch{1.6}{
Median run times, in minutes, for the analysis of simulation study data from the six scenarios. 
Eagle is compared against five other multi-locus programs/packages (A) and two single-locus programs (B). 
The x- and y-axes are on a log scale for improved aesthetics. Eagle has the lowest run-times of the multi-locus 
programs/packages, sometimes by orders of magnitude. Eagle can even produce results faster than single-locus programs. 
The median run times for the programs/packages across the scenarios are given in the table (C). The entries in a blue font 
correspond to the lowest run-time for a scenario. 
 FaST-LMM$^{all}$ is where calculation of the similarity matrix is based on all the SNP data.  
 FaST-LMM$^{few}$ is where calculation  of the similarity matrix is based on a subset of the SNP data. }
 }
\vspace{1cm}
%% plot formed via PowerPoint - ~/AM-Paper/Figure_power.pptx
%% script on mac
%% ~/AM-Paper/AM+/Plots_for_Paper/time_and_memory_plot_for_paper.R
\label{fig_time}
\centering
    \includegraphics[width=1.2\textwidth,natwidth=610,natheight=642]{Figure2_time.jpg}
\end{figure}

@@

We were interested in comparing results from Eagle with those from single-locus association mapping for a real data set.
 We chose to focus on data from a large outbred mouse study \citep{nicod2016genome}. This study was unusual in that it collected and analysed SNP dosages (continuous values from zero to one of expected allele counts)  instead of the more common SNP genotypes. Analyses based on dosages rather than discrete genotypes have been shown to have greater power for the detection of genes that are influencing a trait  \citep{zheng2011comparison}. By converting the dosages into genotypes and analysing the data with the single-locus program FaST-LMM, we obtained a subset of those findings reported in the original study. We then analysed the data with Eagle. Due to Eagle's increased power, we found SNP-trait associations not found with  FaST-LMM. However, we were 
 able to confirm the validity of these new findings as they matched what was found in the original study. Having the ability to confirm new findings  in a real study was 
 one of the primary motivators for choosing these data for analysis.

We repeated the single-locus analyses as first performed \citep{nicod2016genome}, except 
that we focused on autosomal SNPs and our analyses were based on SNP genotypes  rather than SNP dosages. 
In the original analysis, a genome-wide threshold that gave a false discovery rate of 5\%, was found via permutation. We followed the 
same empirical procedure but increased the number of permutations from 100 to 500 for more accurate thresholds.

We ran Eagle in three ways. 
Eagle chooses the best model via the extended Bayesian information criteria (extBIC) \citep{chen2008extended}. 
  The conservativeness of the extBIC can be adjusted by a single regularisation parameter  $\gamma$ that ranges from zero to one. 
In the simulation study, this parameter was set to one, its most conservative and default setting. The mouse data were also analysed 
under this setting (Eagle$^{default}$).  An alternate  \citep{chen2008extended} , less conservative way of setting $\gamma$ is to  let $\gamma = 1 - \frac{1}{(2\kappa)}$ with $\kappa=\frac{log(L)}{log(n_g)}$ where 
 $L$ is the number of loci that span the genome, and $n_g$ is the number of individuals/groups/lines/strains in the study
  (Eagle$^{alt}$). However, our preferred way is to set the $\gamma$ parameter for each trait via permutation (Eagle$^{optimal}$).  We used 100 permutations to set $\gamma$ to give a false  positive rate of 5\%.  This only took six times as long as a single analysis of the data. This is because the marker data need only be read  once, 
 and only the trait data changes across permutations leading to other computational efficiencies.  
  This permutation method has been implemented within the
 Eagle package.

@@

\begin{figure}
\caption{
\setstretch{1.6}{
Genome-wide association mapping results from analyses of the mouse data for the single-locus method FaST-LMM and the 
multi-locus method
Eagle. Eagle was run under three settings; its default setting (Eagle$^{default}$),  an alternate less conservative setting based on 
the number of SNPs and sample size (Eagle$^{alt}$), and where the model selection had been optimised for a false positive rate 
of 5\% (Eagle$^{optimal}$).  The number of SNP-trait associations found are reported in the cells. }
}
\label{figmouse}
\begin{center}
%% obtained from bracewell by running ResultsV2.R from /data/geo047/MWAM/Mouse/OutBred/RScripts/
\includegraphics[width=15cm, height=15cm]{mouseresults.eps}
\end{center}

@@

\section{Discussion/Conclusion}
Eagle is a new linear mixed model based method (and R package) for multi-locus association mapping. It advances the state of association mapping in several ways. 
First, its computational footprint is much smaller than other multi-locus implementations. Eagle makes multi-locus analysis 
practical, even when the datasets are large. Second, the results from
 Eagle are immediately interpretable. They are the set of SNPs in strongest association with the trait where 
each SNP identifies a separate genomic region of interest. Third, it treats association mapping as a model selection problem, avoiding 
multiple testing issues and the need for significance thresholds. 
 As we saw in the simulation study, Eagle has considerably higher power than single-locus methods but is comparable in run time.
Also, when analysing the mouse data, Eagle found more than double the SNP-trait associations than 
with single-locus association mapping, the method of choice. Furthermore, these extra findings were all true.

Eagle outperformed the other multi-locus methods in our simulation study. However, we are cognisant of the fact  that we made several implementation 
choices that impact our conclusions.  For instance, we chose to calculate the significance of the 
SNP effects from bigRR, LMM-Lasso, and glmnet via stability selection.  Permutation and its variants \citep{ browning2008presto,pahl2010permory} are also equally valid empirical approaches. Stability selection though has the advantage of being based on repeated sampling of only a proportion (50\% in our case) of the 
data. Also, when analysing the (sub)samples, it was not necessary to calculate the entire solution path for a method. 
 Instead,  analyses are 
performed for a fixed value of the regularisation parameter, greatly reducing the amount of computation required. For r2VIM, an R package 
implementing random forests, we had to decide on the  minimum size of a terminal node, the number of trees, and number of potential variables. 
The setting of these parameters greatly affects performance.  We  acknowledge that in the hands of an expert, 
r2VIM could be fine-tuned for a better balance of computational and statistical performance. 
However, we would like to think that the parameter settings we used are sensible 
since they match the values in the original r2VIM publication \citep{szymczak2016r2vim}.
